---
title: "Metapopulations"
date: BIO605 FA20
output:
  revealjs::revealjs_presentation:
    self_contained: false
    highlight: haddock
    css: style_slide.css
    theme: white
    reveal_plugins: ["menu", "chalkboard"]
    reveal_options:
      slideNumber: true
      chalkboard:
        theme: whiteboard
        toggleNotesButton: true
      menu:
        numbers: true
---

```{r setup, include = F}

knitr::opts_chunk$set(echo = F, fig.width = 5, fig.height = 4, fig.align = "center")
CEX <- LWD <- 2

```

# Population ecology

## Population ecology

What do population ecologists study?

**Population dynamics** --- how population size/density changes over time in a single habitat

**Distributions** --- how populations are distributed over space

**Landscape-level dynamics** --- how populations are maintained at the landscape level (**metapopulation**)

## Examples

<div id="left">
**Population dynamics**

```{r, echo=FALSE}

set.seed(1)
n <- 20; x <- 1:n; y <- rpois(n, 30)
par(mar = c(5,5,3,2) )
plot(y ~ x,
     ylab = "Population density", xlab = "Year", cex.lab = CEX,
     type = "o", cex = CEX, lwd = LWD, col = "steelblue", axes = F)
box(bty = "l")

```

birth & death\
(immigration, emigration)
</div>

<div id="right">
**Spatial distribution**

```{r, echo = FALSE}

set.seed(3)
par(mar = c(5,5,3,2) )
n <- 20; x <- runif(n, 0, 100); y <- runif(n, 0, 100)
plot(x ~ y, pch = 21, cex = CEX, bg = grey(rbinom(n, 1, p = 0.5)),
     axes = F, xlab = "X-coordinate", ylab = "Y-coordinate", cex.lab = CEX)
box(bty = "l")

```

local environments, dispersal
</div>

# Classical metapopulation

## What is a metapopulation?

<span style="color:steelblue">**Population of populations**</span>

- Populations are connected by dispersal
- Does this make difference?

## Recall: local population

Local population dynamics with no dispersal is driven by birth $b$ and death $d$ processes

$$
\frac{dN}{dt} = (b - d)N
$$

Once the population goes extinct, **no chance to re-establish a new population**

## Immigration

Immigration can lead to a colonization into a new habitat

```{r, fig.width = 5}

plot(0, type = "n", xlim = c(0,1), ylim = c(0,1),
     axes = F, ann = F)
points(x = c(0.2, 0.8), y = c(0.5, 0.5), cex = 10,
       pch = 21, bg = c(grey(1), grey(0,0.5)), col = grey(0,0.5))
arrows(x0 = 0.8, y0 = 0.5, x1 = 0.2, col = grey(0,0.5) )
mtext("Colonization", cex = CEX)

```

## Metapopulation theory

**Classical metapopulation theory** focuses on dynamics at a landscape level

- Local population dynamics are not explicitly modeled
  - either `occupied` or `unoccupied`
  - no information on population size

- Metapopulation persistence is the primary interest
  - how patch occupancy changes over time
  - key: colonization $m$* and extinction rates $e$*
  
<small>*Parameter notations may differ among studies</small>

## Levins metapopulation model

The Levins Metapopulation Model:

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

- $p$: proportion of patches occupied
- $m$: colonization rate (per patch)
- $e$: extinction rate (per patch)

<small>
Levins (1969) Some demographic and genetic consequences of environmental heterogeneity for biological control. Bulletin of the Entomological Society of America. 15: 237-240
</small>

## Colonization

The Levins Metapopulation Model:

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

The term $mp(1-p)$ determines the rate of increase in occupancy

  - immigrants come from occupied patches ($p$)
  - colonization occurs only at unoccupied patches
  - multiply $m$ to express random colonization


## Extinction

The Levins Metapopulation Model:

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

The term $ep$ determines the rate of decrease in occupancy

  - extinction occurs only at occupied patches ($p$)
  - multiply $e$ to express random extinction 


## R exercise: colonization

Draw how colonization rate changes over $p$

Create $p$ and $m$ (set 1.0 for an example)

```{r, echo = TRUE}

p <- seq(from = 0, to = 1, length = 100)
m <- 1

```

## R exercise: colonization

Draw how colonization rate changes over $p$

Plot the relationship b/w $p$ and $mp(1-p)$

```{r, fig.height = 3, fig.width = 5, echo = TRUE, eval = F}

M <- m*p*(1-p)
plot(M ~ p, type = "l")

```

## R exercise: colonization

Draw how colonization rate changes over $p$

Plot the relationship b/w $p$ and $mp(1-p)$

```{r, fig.height = 3, fig.width = 5, echo = TRUE}

M <- m*p*(1-p)
plot(M ~ p, type = "l")

```

## R exercise: extinction

Draw how extinction rate changes over $p$

Create $e$ (set 0.2 for an example)

```{r, echo = TRUE}

e <- 0.2

```

## R exercise: extinction

Draw how extinction rate changes over $p$

Plot the relationship b/w $p$ and $ep$

```{r, fig.height = 3, echo = TRUE, eval = F}

E <- e*p
plot(E ~ p, type = "l")

```

## R exercise: extinction

Draw how extinction rate changes over $p$

Plot the relationship b/w $p$ and $ep$

```{r, fig.height = 3, echo = TRUE, eval = T}

E <- e*p
plot(E ~ p, type = "l")

```

## R exercise: m & e

Draw how colonization & extinction rates change over $p$

```{r, fig.height = 3, echo = TRUE}

plot(M ~ p, type = "l")
lines(E ~ p, col = "red")

```


## Equilibrium occupancy

```{r}

par(mar = c(5,5,4,3))
plot(M ~ p, type = "l", axes = F,
     ylab = "Colonization or extinction rate", xlab = "p")
lines(E ~ p, col = "red")
box(bty = "l")
axis(1); axis(2, las = 2)

```

- With $p$ below the intersection, occupancy ___
- With $p$ above the intersection, occupancy ___


## Equilibrium occupancy

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

```{r}

par(mar = c(5,5,4,3))
plot(I(M-E) ~ p, type = "l", axes = F,
     ylab = "dp/dt", xlab = "p")
points(I(1-e/m), 0, cex = CEX)
abline(h = 0, col = grey(0, 0.2) )
box(bty = "l")
axis(1); axis(2, las = 2)

```


## Equilibrium occupancy

How to get equilibrium occupancy?


## Equilibrium occupancy

Solve the following eq. about $p$

$$
\begin{align}
\frac{dp}{dt} &= cp(1-p) - ep = 0\\
p^* &= ...\\
\end{align}
$$


## Equilibrium occupancy

Solve the following eq. about $p$

$$
p^* = 1 - \frac{e}{m}
$$


## Condition for persistence

For a metapopulation to persist, $p^*$ > 0

$$
p^* = 1 - \frac{e}{m} > 0
$$


## Condition for persistence

For a metapopulation to persist, $p^*$ must exceed zero, meaning

$$
m > e
$$

Colonization rate $m$ must exceed extinction rate $e$


## Link to logistic model

The Levins Metapopulation Model:

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

can be transformed to:

$$
\frac{dp}{dt} = (m-e)p(1-\frac{p}{1-\frac{e}{m}})
$$


## Link to logistic model

Express differently

$$
\frac{dp}{dt} = (m-e)p(1-\frac{p}{1-\frac{e}{m}})
$$

Let (1) $p = N$, (2) $m-e = r$, and (3) $1-\frac{e}{m} = K$

$$
\frac{dN}{dt} = rN(1-\frac{N}{K})
$$

Thus, the Levins metapopulation model can be seen as a logistic model at a different spatial scale

<small>NOTE: not identical as $0 \le p \le 1$</small>


## Assumptions

Classical metapopulation assumes

- No variation in population size
- Immigration from within-metapopulation sources
- Random colonization & extinction
- Dispersal do NOT affect population dynamics



# Mainland-island metapopulation

## Mainland-island metapopulation model

One of the assumptions in the Levins metapopulation model is equal population size

The mainland-island metapopulation model is on the other extreme

## Levins vs. mainland-island

<div id="left">
```{r}
set.seed(1)

x <- runif(10, -1, 1)
y <- runif(10, -1, 1)

par(mar = c(0,0,3,0) )
plot(y ~ x, cex = 3,
     xlim = c(-1.2,1.2), ylim = c(-1.2,1.2),
     ann = F, axes = F,
     pch = 21, bg = grey(rbinom(10, 1, 0.5), 0.5) )
mtext("Levins model", cex = 2)

```
</div>

<div id="right">
```{r}
set.seed(1)

x <- runif(10, -1, 1)
y <- runif(10, -1, 1)

par(mar = c(0,0,3,0) )
plot(y ~ x, cex = 3,
     xlim = c(-3,1.2), ylim = c(-1.2,1.2),
     ann = F, axes = F,
     pch = 21, bg = grey(rbinom(10, 1, 0.5), 0.5) )
points(-2.2, 0, cex = 18, pch = 21, bg = grey(0, 0.5))
mtext("Mainland-island model", cex = 2)

```
</div>

## Assumptions

- One LARGE population (the mainland) that never goes extinct
- Immigration from the mainland to islands only
- Random extinctions on islands

## The formula

The mainland-island metapopulation model:

$$
\frac{dp}{dt} = m(1-p) - ep
$$

The Levins metapopulation model:

$$
\frac{dp}{dt} = mp(1-p) - ep
$$

## R exercise: colonization

Draw how colonization/extinction rate changes over $p$

Create $p$, $m$ and $e$

```{r, echo = T}
p <- seq(from = 0, to = 1, length = 100)
m <- e <- 1
```

## R exercise: colonization

Draw how colonization/extinction rate changes over $p$

Create $m(1-p)$ and $ep$

```{r, echo = T}
M <- m*(1-p)
E <- e*p
```

## R exercise: colonization

Plot $m(1-p)$ and $ep$

```{r, echo = T}
plot(M ~ p, type = "l", ylab = "Colonization or extinction rate")
lines(E ~ p, col = "red")
```

## Equilibrium

The mainland-island metapopulation model:

$$
\frac{dp}{dt} = m(1-p) - ep
$$

The equilibrium occupancy $p^*$ is

$$
p^* = ...
$$

## Equilibrium

$$
p^* = \frac{m}{m+e}
$$

```{r}
m <- e <- 1
p <- seq(0, 1, length = 100)
M <- m*(1-p)
E <- e*p

plot(M ~ p, type = "l", axes = F, ylab = "Colonization or Extinction")
lines(E ~ p, col = "red")
box(bty = "l")
axis(1); axis(2, las = 2)

```

## Condition for persistence

For a metapopulation to persist, $p^* > 0$

$\frac{m}{m+e} > 0$, i.e., $m > 0$


## Condition for persistence

When $m > 0$, there is a persistent supply of immigrants from the mainland 

- the metapopulation never goes extinct
- higher extinction rate $e$ cannot lead to the metapopulation extinction


# Field study

## More complexity

The metapopulation models are clearly oversimplified

In particular...

- Random extinction
- Random colonization


## Assumptions

**Random extinction**

Population size varies in nature and influences local extinction risk

**Random colonization**

Organisms have limited dispersal capability


## Proxies

Focus on a single habitat patch and think colonization into and extinction of the patch

```{r}
set.seed(2)
x <- runif(10)
y <- runif(10)
par(mar = rep(0,4) )
plot(y ~ x, cex = runif(10, 0.1, 12), ann = F, axes = F,
     xlim = c(-0.2,1.2), ylim = c(-0.2,1.2) )
```


## Proxies

**Habitat size**

Extinction probability decreases with increasing habitat size

```{r}
set.seed(2)
n <- 10
x <- runif(n)
y <- runif(n)
cex <- runif(n, 0.1, 12)
bg <- 1 - scale(cex, center = min(cex), scale = max(cex) - min(cex))

par(mar = rep(0,4) )
plot(y ~ x, cex = cex, ann = F, axes = F,
     xlim = c(-0.2,1.2), ylim = c(-0.2,1.2),
     pch = 21, bg = grey(bg) )
```


## Rationale: habitat size

In a large habitat...

- population size is larger
- more refuge


## Proxies

**Isolation**

Colonization probability decreases with increasing isolation

```{r}
set.seed(2)
n <- 10
x <- runif(n)
y <- runif(n)
alpha <- 10

D <- as.matrix(dist(cbind(x,y), diag = T, upper = T))
C <- colSums(exp(-alpha*D))

cex <- runif(n, 0.1, 12)
bg <- 1 - scale(C, center = min(C), max(C) - min(C) )

par(mar = rep(0,4) )
plot(y ~ x, cex = cex, ann = F, axes = F,
     xlim = c(-0.2,1.2), ylim = c(-0.2,1.2),
     pch = 21, bg = grey(bg) )
```


## Rationale: isolation

In a isolated habitat,

- dispersal is more costly & risky
- dispersing individuals are less likely to arrive

Measures of isolation

- distance to the nearest neighbor
- distance weighted measure


## Empirical approach

When studying **real** organisms...

- survey presence/absence of the species $y_i$ at each habitat
- relate to habitat size and isolation (and other factors)

$$
y_i = f(habitat~size, isolation)
$$


## Reality

More reality might be needed for empirical studies

> - matrix permeability
> - wind
> - current
> - species interactions



## Gap

**Theory**: metapopulation-level occupancy

**Empirical**: patch-level occupancy

## Testing metapopulation theory

Patch-level studies can still provide insights

However, our primary interest is **the metapopulation persistence**, not persistence of local populations

## Approach to studying metapopulations

**Metapopulation replicates** are needed to study the drivers of metapopulation persistence

or

**Scaling up** with simulation approaches with parameters estimated with patch-level studies





