---
title: "Competition"
date: BIO605 FA20
output:
  revealjs::revealjs_presentation:
    self_contained: false
    highlight: haddock
    css: style_slide.css
    theme: white
    reveal_plugins: ["menu", "chalkboard"]
    reveal_options:
      slideNumber: true
      chalkboard:
        theme: whiteboard
        toggleNotesButton: true
      menu:
        numbers: true
---


```{r setup, include = F}

knitr::opts_chunk$set(echo = F, fig.width = 5, fig.height = 4, fig.align = "center")
CEX <- LWD <- 2

```

# Community Ecology

## What is a community?

A group of species that interact in a given space

  - Communities have structure, including **species richness**, **relative abundance** etc.
  - Different types of interactions
  - Why is it important to study interactions in a community?

## Interctions matter

Species interactions can be a determinant of:

  - persistence of constituent species
  - stability of a community
  - ecosystem functions
  
## Types of interactions

List species interactions

> - <span style="color:steelblue;">Competition</span>
> - Predation
> - Mutualism
> - Parasitism
> - Commensalism
> - and more...

## Competition

Competition is considered to be a dominant (perhaps THE dominant) type of species interactions in a community*.

<small>*I doubt it</small>

**Contest competition**

One or few individuals dominate resources

**Scramble competition**

Depletes resources and all competitors are affected


# Competition models

## Recall: logistic model

One species model (Logistic model):

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K}) &&\text{Logistic model}\\
\end{align}
$$


## Include competition

Add $N_2$ to the equation:

$$
\begin{align}
\frac{dN_1}{dt} &= r_1N_1(1-\frac{N_1+N_2}{K_1})\\
\end{align}
$$

How does $N_2$ impact $\frac{dN_1}{dt}$?

## R exercise: set parameters

Visualize how $N_2$ impacts $\frac{dN_1}{dt}$

Set $r$, $K_1$, $N_1$, and $N_2$

<small>$N_2$ is set to be $0$ for reference</small>

```{r, echo = TRUE}
r1 <- 1.5
K1 <- 100
N1 <- seq(from = 0, to = 150, length = 100)
N2 <- 0 # set 0 as a reference
```

## R exercise: equation

Write equation

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+N_2}{K_1})\\
$$

```{r, echo = TRUE}

dN1_0 <- r1*N1*(1 - (N1 + N2)/K1)

```


## R exercise: visualize

```{r, echo = TRUE}

plot(dN1_0 ~ N1, type = "l", ylab = "dN1/dt") # N2 = 0
abline(h = 0, col = grey(0, 0.2))
points(100, 0, cex = CEX)
```


## R exercise: visualize

```{r, echo = TRUE, eval = F}

dN1_10 <- r1*N1*(1 - (N1 + 10)/K1) # N2 = 10
plot(dN1_0 ~ N1, type = "l", ylab = "dN1/dt")
abline(h = 0, col = grey(0, 0.2))
lines(dN1_10 ~ N1, col = "blue")

```

## R exercise: visualize

```{r, echo = TRUE, eval = T}

dN1_10 <- r1*N1*(1 - (N1 + 10)/K1) # N2 = 10
plot(dN1_0 ~ N1, type = "l", ylab = "dN1/dt")
abline(h = 0, col = grey(0, 0.2))
lines(dN1_10 ~ N1, col = "blue")

```

## R exercise: visualize

```{r, echo = TRUE, eval= F}

dN1_50 <- r1*N1*(1 - (N1 + 50)/K1) # N2 = 50
plot(dN1_0 ~ N1, type = "l", ylab = "dN1/dt")
abline(h = 0, col = grey(0, 0.2))
lines(dN1_10 ~ N1, col = "blue")
lines(dN1_50 ~ N1, col = "red")

```

## R exercise: visualize

```{r, echo = TRUE, eval= T}

dN1_50 <- r1*N1*(1 - (N1 + 50)/K1) # N2 = 50
plot(dN1_0 ~ N1, type = "l", ylab = "dN1/dt")
abline(h = 0, col = grey(0, 0.2))
lines(dN1_10 ~ N1, col = "blue")
lines(dN1_50 ~ N1, col = "red")

```

## Impact of competition

Logistic model + species 2 ($N_2$)

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+N_2}{K_1})\\
$$

- As $N_2$ increases, the population growth of $N_1$ becomes negative more quickly
- Population size of species 1 reaches its equilibrium below $K_1$, the carrying capacity in the absence of species 2


## Assumptions

Logistic model + species 2 ($N_2$)

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+N_2}{K_1})\\
$$

The above model assumes per-capita impacts are equal b/w species 1 and 2


## Competition coefficient

Logistic model + species 2 ($N_2$)

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+ \alpha N_2}{K_1})\\
$$

Multiply $\alpha$ to model different impacts of species 2

$\alpha$ is referred to as the **competition coefficient**

- $\alpha < 0$ ... effect of species 2 is less than species 1
- $\alpha > 0$ ... effect of species 2 is greater than species 1


## Assumptions

Logistic model + species 2 ($N_2$)

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+ \alpha N_2}{K_1})\\
$$

The above model assumes

- $N_2$ is constant
- $N_1$ has no impacts on $N_2$


## Lotka-Volterra model

Lotka-Volterra model

$$
\frac{dN_1}{dt} = r_1N_1(1-\frac{N_1+ \alpha_{12} N_2}{K_1})\\
\frac{dN_2}{dt} = r_2N_2(1-\frac{N_2+ \alpha_{21} N_1}{K_2})\\
$$

The above equation models dynamic interactions of the two competing species


## Lotka-Volterra model

Lotka-Volterra model (different form)

$$
\frac{dN_1}{dt} = N_1(r_1-\beta_1 N_1 - \gamma_{12}N_2)\\
\frac{dN_2}{dt} = N_2(r_2-\beta_2 N_2 - \gamma_{21}N_1)\\
$$

- $\beta_i$ **Intraspecific** competition coefficient
- $\gamma_{ji}$ **Interspecific** competition coefficient


## Relate parameters

$$
\begin{align}
\frac{dN_1}{dt} &= r_1N_1(1-\frac{N_1+ \alpha_{12} N_2}{K_1})\\
                &= N_1(r_1 - \frac{r_1}{K_1}N_1 - \frac{r_1 \alpha_{12}}{K_1} N_2)\\
                &= N_1(r_1 - \beta_1 N_1 - \beta_1 \alpha_{12} N_2)\\
                &=  N_1(r_1 - \beta_1 N_1 - \gamma_{12} N_2)\\
\end{align}
$$

where $\beta_1 = \frac{r_1}{K_1}$ and $\alpha_{12} = \frac{\gamma_{12}}{\beta_1}$

$\alpha_{12}$ is the **ratio of inta- and interspecific competition coefficitents**


# Predict consequences of competition

## Model prediction

<div id="left">

```{r}
par(mar= c(0,0,0,0), oma = c(0,0,0,0))
plot(0, type = "n", xlim = c(0,1), ylim = c(0,1), ann = F, axes = F)
text(0.5, 0.8, "Mechanism", cex = CEX)
text(0.5, 0.2, "Pattern\n(observation)", cex = CEX)
arrows(x0 = 0.7, y0 = 0.3, y1 = 0.7, col = "gray", lwd = LWD); text(0.3, 0.5, "Theoretical\nmodel", cex = CEX)
arrows(x0 = 0.3, y0 = 0.7, y1 = 0.3, col = "gray", lwd = LWD); text(0.7, 0.5, "Statistical\nmodel", cex = CEX)

```

</div>

<div id="right">

**Theory**

- Generate predictions with given mechanisms

</div>


## Model prediction

Much of interests in community ecology is the coexistence of competiting species

What does the Lotka-Volterra model predict?


## Discrete version

Make the discrete version of the Lotka-Volterra model

In analogy of trasformation from the logistic to Beverton-Holt model

$$
N_{1,t+1} = (\frac{\lambda_1}{1 + \beta_1 N_{1,t} + \gamma_{12}N_{2,t}}) N_{1,t}\\
N_{2,t+1} = (\frac{\lambda_2}{1 + \beta_2 N_{2,t} + \gamma_{21}N_{1,t}}) N_{2,t}
$$


## R exercise: set parameters

Set $\lambda$, $\beta_i$, and $\gamma_{ji}$

```{r,echo=TRUE}

# Set equal values of lambda for species 1 and 2
lambda <- 3

b1 <- 0.1   # Intra-specific competition
g21 <- 0.05 # Impact of sp1 on sp2

b2 <- 0.3   # Intra-specific competition
g12 <- 0.15 # Impact of sp2 on sp1

```


## R exercise: initial abundance

Set initial abundance `N1[1]` and `N2[1]`

```{r,echo=TRUE}

N1 <- N2 <- NULL # create "NULL" objects
N1[1] <- 10  # initial abundance 10
N2[1] <- 5 # initial abundance 5

```

## R exercise: equation

Write the equations

$$
N_{1,t+1} = (\frac{\lambda_1}{1 + \beta_1 N_{1,t} + \gamma_{12}N_{2,t}}) N_{1,t}\\
N_{2,t+1} = (\frac{\lambda_2}{1 + \beta_2 N_{2,t} + \gamma_{21}N_{1,t}}) N_{2,t}
$$

```{r,echo=TRUE}

for(t in 1:99){ # simulate 100 time steps
  N1[t+1] <- (lambda*N1[t])/(1 + b1*N1[t] + g12*N2[t])
  N2[t+1] <- (lambda*N2[t])/(1 + b2*N2[t] + g21*N1[t])
}

```

## R exercise: print

Check the first 5 time steps

```{r,echo=TRUE}

print(N1[1:5])
print(N2[1:5])

```


## R exercise: visualize

```{r, echo = TRUE, eval= F}

t <- 1:length(N1)

plot(N1 ~ t, type = "l", col = "blue", ylab = "N1 or N2",
     ylim = c(0, max(N1, N2) ) )
lines(N2 ~ t, col = "red")

```

## R exercise: visualize

```{r, echo = TRUE}

t <- 1:length(N1)

plot(N1 ~ t, type = "l", col = "blue", ylab = "N1 or N2",
     ylim = c(0, max(N1, N2) ) )
lines(N2 ~ t, col = "red")

```


## Model prediction

$$
N_{1,t+1} = (\frac{\lambda_1}{1 + \beta_1 N_{1,t} + \gamma_{12}N_{2,t}}) N_{1,t}\\
N_{2,t+1} = (\frac{\lambda_2}{1 + \beta_2 N_{2,t} + \gamma_{21}N_{1,t}}) N_{2,t}
$$

In the previous exercise, we set:

> - $\lambda_1 = \lambda_2$ = `r lambda`
> - $\beta_1$ = `r b1` and $\gamma_{21}$ = `r g21`
> - $\beta_2$ = `r b2` and $\gamma_{12}$ = `r g12`

What does this mean?

## Competition: scenario 1

Try the following conditions

**Scenario 1**

Intraspecific competition $\beta_{i}$ is **stronger** than interspecific competition $\gamma_{ji}$ for both species
($\beta_i > \gamma_{ji}$)


## R exercise: scenario 1

```{r,echo=TRUE}

# Set equal values of lambda for species 1 and 2
lambda <- 3

b1 <- 0.1   # Intra-specific competition
g21 <- 0.05 # Impact of sp1 on sp2

b2 <- 0.3   # Intra-specific competition
g12 <- 0.15 # Impact of sp2 on sp1

# initial abundance
N1 <- N2 <- NULL # create "NULL" objects
N1[1] <- 5  # set initial abundance 5
N2[1] <- 10 # set initial abundance 10

# simulate
for(t in 1:99){ # simulate 100 time steps
  N1[t+1] <- (lambda*N1[t])/(1 + b1*N1[t] + g12*N2[t])
  N2[t+1] <- (lambda*N2[t])/(1 + b2*N2[t] + g21*N1[t])
}

```

## R exercise: scenario 1

```{r, echo = TRUE, fig.height=3, fig.width=5}

t <- 1:length(N1)

plot(N1 ~ t, type = "l", col = "blue", ylab = "N1 or N2",
     ylim = c(0, max(N1,N2) ) )
lines(N2 ~ t, col = "red")

```

## Prediction: scenario 1

When $\beta_i > \gamma_{ji}$,

the **two species coexist** (initial abundance does not matter)

## Competition: scenario 2

Try the following conditions

**Scenario 2**

Intraspecific competition $\beta_{i}$ is **stronger** than interspecific competition $\gamma_{ji}$ for species 1
($\beta_1 > \gamma_{21}$)

&

Intraspecific competition $\beta_{i}$ is **weaker** than interspecific competition $\gamma_{ji}$ for species 2
($\beta_2 < \gamma_{12}$)


## R exercise: scenario 2

Set initial abundance as you like

```{r,echo=TRUE}

# Set equal values of lambda for species 1 and 2
lambda <- 3 

# in sp.1 equation
b1 <- 1.5
g21 <- 1.0

# in sp.2 equation
b2 <- 0.1
g12 <- 0.8

# initial abundance
N1 <- N2 <- NULL # create "NULL" objects
N1[1] <- 10
N2[1] <- 8

# simulate
for(t in 1:99){ # simulate 100 time steps
  N1[t+1] <- (lambda*N1[t])/(1 + b1*N1[t] + g12*N2[t])
  N2[t+1] <- (lambda*N2[t])/(1 + b2*N2[t] + g21*N1[t])
}

```

## R exercise: scenario 2

```{r, echo = TRUE, fig.height=3, fig.width=5}

t <- 1:length(N1)

plot(N1 ~ t, type = "l", col = "blue", ylab = "N1 or N2",
     ylim = c(0, max(N1, N2) ) )
lines(N2 ~ t, col = "red")

```

## Prediction: scenario 2

When $\beta_i > \gamma_{ji} \cap \beta_j < \gamma_{ij}$ (or $\beta_i < \gamma_{ji} \cap \beta_j > \gamma_{ij}$),

the **two species cannot coexist**

- Winner has greater impacts on the competitor than themselves ($\beta_i < \gamma_{ji}$ or $\beta_j < \gamma_{ij}$)
- Loser has greater impacts on themselves than the competitor ($\beta_i > \gamma_{ji}$ or $\beta_j > \gamma_{ij}$)

## Competition: scenario 3

Try the following conditions

**Scenario 3**

Intraspecific competition $\beta_{i}$ is **weaker** than interspecific competition $\gamma_{ji}$ for species 1
($\beta_1 < \gamma_{21}$)

&

Intraspecific competition $\beta_{i}$ is **weaker** than interspecific competition $\gamma_{ji}$ for species 2
($\beta_2 < \gamma_{12}$)


## R exercise: scenario 3

Set initial abundance as you like

```{r,echo=TRUE}

# Set equal values of lambda for species 1 and 2
lambda <- 3 

# in sp.1 equation
b1 <- 0.1
g21 <- 0.2

# in sp.2 equation
b2 <- 0.2
g12 <- 0.4

# initial abundance
N1 <- N2 <- NULL # create "NULL" objects
N1[1] <- 1
N2[1] <- 1.5

# simulate
for(t in 1:99){ # simulate 100 time steps
  N1[t+1] <- (lambda*N1[t])/(1 + b1*N1[t] + g12*N2[t])
  N2[t+1] <- (lambda*N2[t])/(1 + b2*N2[t] + g21*N1[t])
}

```

## R exercise: scenario 3

```{r, echo = TRUE, fig.height=3, fig.width=5}

t <- 1:length(N1)

plot(N1 ~ t, type = "l", col = "blue", ylab = "N1 or N2",
     ylim = c(0, max(N1, N2) ) )
lines(N2 ~ t, col = "red")

```

## Prediction: scenario 3

When $\beta_i < \gamma_{ji}$,\
the **two species cannot coexist** (winner is the species with higher initial abundance)

## Summary

Lotka-Volterra model predicts three consequences

**Stable coexistence**\
$\beta_i > \gamma_{ji}$

**Competitive dominance (exclusion)**\
$\beta_i > \gamma_{ji} \cap \beta_j < \gamma_{ij}$ (or $\beta_i < \gamma_{ji} \cap \beta_j > \gamma_{ij}$)

**Destabilizing competition**\
$\beta_i < \gamma_{ji}$