---
title: "Population Models"
date: BIO605 FA20
output:
  revealjs::revealjs_presentation:
    self_contained: false
    highlight: haddock
    css: style_slide.css
    theme: white
    reveal_plugins: ["menu", "chalkboard"]
    reveal_options:
      slideNumber: true
      chalkboard:
        theme: whiteboard
        toggleNotesButton: true
      menu:
        numbers: true
---

```{r setup, include = F}

knitr::opts_chunk$set(echo = F, fig.width = 5, fig.height = 4)
CEX <- LWD <- 2

```


# Population ecology

## What is a population?

A population is **a group of individuals of the same species** that inhabit a given area

  - Populations have structure, including **density**, **spacing** and **age** distribution
  - Why is it important that the individuals are members of the same species?
  - Why is it important for a population to have a spatial boundary?


## Why is it important?

Population ecology is 

  - important for understanding the **demography** of the species
  - the **basis** of community ecology and other sub-disciplines of ecology
  - important for conservation, especially in the **population viability analysis**
  - useful for the control of **invasive species**

  

## Population ecology

What do population ecologists study?

**Population dynamics** --- how population size/density changes over time in a single habitat

**Distributions** --- how populations are distributed over space

**Landscape-level dynamics** --- how populations are maintained at the landscape level (**metapopulation**)


## Examples

<div id="left">
**Population dynamics**

```{r, echo=FALSE}

set.seed(1)
n <- 20; x <- 1:n; y <- rpois(n, 30)
par(mar = c(5,5,3,2) )
plot(y ~ x,
     ylab = "Population density", xlab = "Year", cex.lab = CEX,
     type = "o", cex = CEX, lwd = LWD, col = "steelblue", axes = F)
box(bty = "l")

```

birth & death\
(immigration, emigration)
</div>

<div id="right">
**Spatial distribution**

```{r, echo = FALSE}

set.seed(3)
par(mar = c(5,5,3,2) )
n <- 20; x <- runif(n, 0, 100); y <- runif(n, 0, 100)
plot(x ~ y, pch = 21, cex = CEX, bg = grey(rbinom(n, 1, p = 0.5)),
     axes = F, xlab = "X-coordinate", ylab = "Y-coordinate", cex.lab = CEX)
box(bty = "l")

```

local environments, dispersal
</div>



# Exponential model

## What determines population size?

 - Suppose you observe a population of hydra
 - Counted birth & death events everyday
 - Estimated birth & death rates


## Math

Initial population size $N_0$, birth rate $b$ & death rate $d$ 

$$
\begin{align}
N_1 &= N_0 + bN_0 - dN_0\\
N_2 &= N_1 + bN_1 - dN_1\\
...
\end{align}
$$

## Math

When $N_0 = 10$, $b=0.8$ and $d=0.2$ 

$$
\begin{align}
N_1 &= N_0 + bN_0 - dN_0\\
    &= 10 + 0.8*10 - 0.2*10\\
    &= 16\\
\end{align}
$$


## Math

Then $N_1 = 16$, $b=0.8$ and $d=0.2$ 

$$
\begin{align}
N_2 &= N_1 + bN_1 - dN_1\\
    &= 16 + 0.8*16 - 0.2*16\\
    &= 25.6\\
\end{align}
$$


## Simple model

Generalize as $N_t$ and modify the equation

<small>
$$
\begin{align}
N_{t+1} &= N_{t} + bN_{t} - dN_{t} &&\text{Generalized}\\
N_{t+1} - N_{t} &= bN_{t} - dN_{t} &&\text{Subtract $N_{t}$ from both sides}\\
N_{t+1} - N_{t} &= (b-d)N_{t} &&\text{Organize the equation}\\
\Delta N_{t} &= rN_{t} &&\text{$\Delta N_t = N_{t+1} - N_{t}$, $r = b-d$}\\
\end{align}
$$
</small>

- $\Delta N_{t}$ represents **the net increase** per unit time
- $r = b-d$ determines the population growth **rate**

## Simple model

<div id="left">
```{r}
N <- matrix(10, nrow = 20, ncol = 3)
N[1,] <- 10; r <- 1:3*0.1
Nt <- 20; t <- 1:Nt
for(i in 1:(Nt-1) ){ N[i+1,] <- N[i,] + r*N[i,] }

par(mar = c(5,5,4,3))
matplot(t, N, type = "o", pch = 21, lty = 1,
        axes = F,
        cex = CEX, lwd = LWD, cex.lab = CEX, cex.axis = CEX,
        ylab = "Population size N", xlab = "t",
        col = rainbow(3) )
box(bty = "l")
```
</div>

<div id="right">
**Model**

$N_{t+1} = N_{t} + rN_{t}$
\
\

**Growth rate** $r$

- $r = `r r[1]`$ <span style="color: red;">red</span> 
- $r = `r r[2]`$ <span style="color: green;">green</span> 
- $r = `r r[3]`$ <span style="color: blue;">blue</span> 
</div>


## Simple model

Geometric population model assumes

- population growth of discrete intervals $\Delta t = 1$
- unit can be one year, day, hour...


## Exponential model

Convert the model to a continuous version

- Birth & death processes can occur continuously
- Take the limit of $\Delta N$
  - $\displaystyle \lim_{\Delta t \to 0} \Delta N$
  - As $\Delta t$ approaches zero, the rate of change become **instantaneous**
  - **Exponential model** $\frac{dN}{dt} = rN$
- Solving $\frac{dN}{dt} = rN$ yields $N_t = e^{rt}N_0$


## Exponential model

<div id="left">
```{r}
N0 <- 10; r <- 1:3*0.1
N <- sapply(1:3, function(i) N0*exp(r[i]*seq(0, 20, length = 100)) ) 

par(mar = c(5,5,4,3))
matplot(N, type = "l", pch = 21, lty = 1,
        axes = F,
        cex = CEX, lwd = LWD, cex.lab = CEX, cex.axis = CEX,
        ylab = "Population size N", xlab = "t",
        col = rainbow(3) )
box(bty = "l")
```
</div>

<div id="right">
**Model**

$$
\begin{aligned}
\frac{dN}{dt} &= rN\\
N_t &= e^{rt}N_0
\end{aligned}
$$

**Population growth rate**

- $r = `r r[1]`$ <span style="color: red;">red</span> 
- $r = `r r[2]`$ <span style="color: green;">green</span> 
- $r = `r r[3]`$ <span style="color: blue;">blue</span> 
</div>

## Exponential model

Recall: $r = b-d$

- What happens if $r = b-d = 0$?\
(i.e., death equals birth)
- What happens if $r = b-d < 0$?\
(i.e., death exceeds birth)



## R exercise: create time data

Create "time" data `t` (x-axis)

- `seq()` is a function to create a vector
- give `from`, `to` and `length`

```{r, echo = TRUE}

t <- seq(from = 0, to = 50, length = 100)

```


## R exercise: check elements

Check elements

```{r, echo = TRUE}
# elements 1 to 5
t[1:5]

# elements 96 to 100
t[96:100]

```


## R exercise: initial condition

Define the initial population size `N0`

```{r, echo = TRUE}

N0 <- 10
print(N0)

```


## R exercise: growth rate

Define the population growth rate `r`

- Set `0` as a reference case

```{r, echo = TRUE}

r0 <- 0
print(r0)

```


## R exercise: equation

Write the equation

- `exp(x)` is $e^x$

```{r, echo = TRUE}

N <- exp(r0*t)*N0

```


## R exercise: visualize

Visualize with `plot()`

```{r, echo = TRUE, fig.width = 3, fig.height = 3, fig.align = "center" }

plot(N ~ t, type = "l", ylim = c(0, 12) )

```


## R exercise: growth rate

Try another parameter

- Set `r1` as `-0.1`

```{r, echo = TRUE, fig.width = 3, fig.height = 3, fig.align = "center" }

r1 <- -0.1
N1 <- exp(r1*t)*N0
plot(N1 ~ t, type = "l", ylim = c(0, 12), col = "red")

```


## R exercise: compare

Compare with `N`

```{r, echo = TRUE, fig.width = 3, fig.height = 3, fig.align = "center" }

plot(N1 ~ t, type = "l", ylim = c(0, 12), col = "red")
lines(N ~ t)

```


## Theory

Recall: what's the role of theory?

<div id="left">

```{r}
par(mar= c(0,0,0,0), oma = c(0,0,0,0))
plot(0, type = "n", xlim = c(0,1), ylim = c(0,1), ann = F, axes = F)
text(0.5, 0.8, "Mechanism", cex = CEX)
text(0.5, 0.2, "Pattern\n(observation)", cex = CEX)
arrows(x0 = 0.7, y0 = 0.3, y1 = 0.7, col = "gray", lwd = LWD); text(0.3, 0.5, "Theoretical\nmodel", cex = CEX)
arrows(x0 = 0.3, y0 = 0.7, y1 = 0.3, col = "gray", lwd = LWD); text(0.7, 0.5, "Statistical\nmodel", cex = CEX)

```

</div>

<div id="right">

**Theory**

- Generate predictions with given mechanisms
- How do population dynamics look like if $r = XX$?
- $r$ is a **parameter**

</div>

## Observation

Recall: what's the role of observation?

<div id="left">

```{r}
par(mar= c(0,0,0,0), oma = c(0,0,0,0))
plot(0, type = "n", xlim = c(0,1), ylim = c(0,1), ann = F, axes = F)
text(0.5, 0.8, "Mechanism", cex = CEX)
text(0.5, 0.2, "Pattern\n(observation)", cex = CEX)
arrows(x0 = 0.7, y0 = 0.3, y1 = 0.7, col = "gray", lwd = LWD); text(0.3, 0.5, "Theoretical\nmodel", cex = CEX)
arrows(x0 = 0.3, y0 = 0.7, y1 = 0.3, col = "gray", lwd = LWD); text(0.7, 0.5, "Statistical\nmodel", cex = CEX)

```

</div>

<div id="right">
**Observation**

- Infer mechanisms (parameters)
- $N_0 = 10$ and $N_{50} = 300$
- **Assume** exponential model
  - $N_t = e^{rt}N_0$
  - $N_{50} = e^{50r}10$?
  - $r = ??$
\
\
\
</div>

<small>
*Note: in practice, parameter inference is much more complex to account for sampling uncertainty
</small>

# Logistic model

## Exponential model

In the exponential model of population growth

- no resource limitation assumed
- a population grows infinitely
- in nature, however, resources are limited

<small> *Exponential model is appropriate for describing dynamics of a newly established population </small>


## Exponential to logistic

Instantaneous population growth $\frac{dN}{dt}$ is

$$
\begin{align}
\frac{dN}{dt} &= rN &&\text{Exponential model}\\
\frac{dN}{dt} &= rN(1-\frac{N}{K}) &&\text{Logistic model}\\
\end{align}
$$

## Logistic model: parameter

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K}) &&\text{Logistic model}\\
\end{align}
$$

- $r$ is the **intrinsic rate** of population growth
- $K$ is the **carrying capacity**


## Logistic model: what if

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K}) &&\text{Logistic model}\\
\end{align}
$$

**Questions**

- If $N < K$, $\frac{dN}{dt}$
- If $N = N$, $\frac{dN}{dt}$
- If $N > N$, $\frac{dN}{dt}$


## Logistic model: visualize

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K})\\
\end{align}
$$

```{r, fig.align="center"}
ylab <- expression(dN/dt)
fun <- function(r,x,K){r*x*(1-x/K)}

x <- 0:120; K <- 100; r <- 1
par(mar=c(5,5,3,2))
plot(fun(r,x,K) ~ x, type = "l", axes = F,
     ylab = ylab, xlab = "N", cex.lab = CEX)
abline(h = 0, col=grey(0.5), lty = 2)
box(bty = "l")
points(K, 0, pch = 21, cex = CEX)
axis(2, at = 0)

```


## Logistic model: competition

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K}) &&\text{Logistic model}\\
\end{align}
$$

The term $1-\frac{N}{K}$

- is a decreasing function of $N$
- expresses **density dependence**
- involves density dependent birth & death

## Logistic model: solve

Solve the equation

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K})\\
N_t &= \frac{K}{1+(\frac{K-N_0}{N_0})e^{-rt}}
\end{align}
$$

## R exercise: define parameters

**Model**

$$
\begin{align}
N_t &= \frac{K}{1+(\frac{K-N_0}{N_0})e^{-rt}}
\end{align}
$$

Create time: `t={0...50}`

Define parameters: `r=1`, `K=200`, `N0=10`

```{r, echo = TRUE}
t <- seq(0, 50, length = 100)
r <- 1
K <- 200
N0 <- 10
```


## R exercise: write the eq.

**Model**

$$
\begin{align}
N_t &= \frac{K}{1+(\frac{K-N_0}{N_0})e^{-rt}}
\end{align}
$$

Write in two lines to avoid errors

```{r, echo = TRUE}

C <- (K - N0)/N0
N <- K/(1 + C*exp(-r*t) )

```

## R exercise: visualize

```{r, echo = TRUE, fig.align = "center" }
plot(N ~ t, type = "l")
```


## R exercise: visualize

**Model**

$$
\begin{align}
N_t &= \frac{K}{1+(\frac{K-N_0}{N_0})e^{-rt}}
\end{align}
$$

Make predictinos under the following scenarios

- What if `r=0.1`, `K=200`, `N0=10` (store as `N1`)
- What if `r=1`, `K=100`, `N0=200` (store as `N2`)
- Plot model predictions on a single figure


## Logistic model: +alpha

Modify the equation to facilitate your understanding

$$
\begin{align}
\frac{dN}{dt} &= rN(1-\frac{N}{K})\\
\frac{dN}{dt} &= N(r-\frac{rN}{K})\\
\frac{dN}{dt} &= (r-\beta N)N &&\text{Use $\beta = \frac{r}{K}$}\\
\end{align}
$$




# Discrete model

## Continuous observation...?

- Exponential & logistic models are **continuous**
- Can you observe population size continuously? No way
- **Discrete models**
- How continuous and discrete models are related?

## Multiplicative expression

- Suppose you made observations at year 0, 1, 2,...
- Simplest expession would be $N_1 = \lambda N_0$
- Population growth $\lambda = \frac{N_1}{N_0}$

## Multiplicative expression

Example

$$
\begin{align}
N_1 &= \lambda N_0\\
N_2 &= \lambda N_1\\
N_3 &= \lambda N_2
\end{align}
$$


## Multiplicative expression

Express differently

$$
\begin{align}
N_3 &= \lambda N_2\\
    &= \lambda * \lambda N_1\\
    &= \lambda * \lambda * \lambda N_0\\
    &= \lambda^3 N_0\\
\end{align}
$$


## Geometric model

Generalize - **Geometric** model

$$
\begin{align}
N_{t+1} &= \lambda N_t &&\text{relate $N_{t+1}$ to $N_t$}\\
N_t &= \lambda^t N_0 &&\text{relate $N_{t}$ to $N_0$}\\
\end{align}
$$


## Geometric model: compare

Compare

$$
\begin{align}
N_t &= \lambda^t N_0 &&\text{Geometric}\\
N_t &=e^{rt} N_0 &&\text{Exponential}\\
\end{align}
$$

Let $e^r$ be $\lambda$...the two models become identical!

**Geometric**\
a population grows when $\lambda > 1$ (i.e. $r > 0$)

**Exponential**\
a population grows when $r > 0$ (i.e. $\lambda > 1$)

## Beverton-Holt model

**Beverton-Holt model**\
Include density dependence ($\beta = \frac{\lambda-1}{K}$)

$$
\begin{align}
N_{t+1} &= \frac{\lambda N_t}{1 + \beta N_t} &&\text{relate $N_{t+1}$ to $N_{t}$}\\
\frac{N_{t+1}}{N_t} &= \lambda_t = \frac{\lambda}{1 + \beta N_t}
\end{align}
$$

- When $N_t = 0$, $\lambda_t$...
- When $N_t = K$, $\lambda_t$...
- When $N_t > K$, $\lambda_t$...

## Beverton-Holt model: solve

**Beverton-Holt model**\
Solve the equation

$$
\begin{align}
N_{t+1} &= \frac{\lambda N_t}{1 + \beta N_t} &&\text{relate $N_{t+1}$ to $N_{t}$}\\
N_{t} &= \frac{K}{1+(\frac{K-N_0}{N_0})\lambda^{-t}} &&\text{relate $N_{t}$ to $N_{0}$}\\
\end{align}
$$


## Beverton-Holt model: compare

Compare

$$
\begin{align}
N_{t} &= \frac{K}{1+(\frac{K-N_0}{N_0})\lambda^{-t}} &&\text{Beverton-Holt}\\
N_t &= \frac{K}{1+(\frac{K-N_0}{N_0})e^{-rt}} &&\text{Logistic}\\
\end{align}
$$

Let $e^r$ be $\lambda$...the two models become identical!


# Summary

## Model types

**Continuous** models are used in pure theoretical research

**Discrete** models can be used in both theoretical and statistical analysis

<div id="left">
**No resource limitation**

- Exponential (continuous)
- Geometric (discrete)

</div>

<div id="right">
**Resource limitation**

- Logistic (continuous)
- Beverton-Holt (discrete)

</div>

## Other models

There are (many) other models

but most of them are a **modification** of these basic models